---
title: "About IODS final assignment"
author: "Abushahba"
date: "December 13, 2017"
output: html_document
---

# **Insights on human dataset**

## *Introduction*

This part of my final assignment will focus on analysing a data set named *humanW*, it originates from the United Nations Development Programme. See [their data page](http://hdr.undp.org/en/content/human-development-index-hdi) for more information. For a nice overview see also [the calculating the human development indices pdf](http://hdr.undp.org/sites/default/files/hdr2015_technical_notes.pdf). Most of the variable names in the data have been shortened and two new variables have been computed. See the [meta file](https://raw.githubusercontent.com/TuomoNieminen/Helsinki-Open-Data-Science/master/datasets/human_meta.txt). For the modified data please check our renamed variables and [wrangled raw data](https://github.com/Abushahba/IODS-final/blob/master/data/humanW.csv), as well as [R script for data wrangling](https://github.com/Abushahba/IODS-final/blob/master/data/create_human.R).

Briefly, the data combines several indicators from most countries in the world that can be broadly classified into those related to **Health and knowledge** and others related to **Empowerment**.

## *Characteristics of the dataset*

First, we will read our working dataset, inspect its structure and dimensions.

```{r}
#read saved modified dataset into an object
humanW <- read.csv(file = "C:/Users/ahmed/Desktop/IODS-final-master/IODS-final/data/humanW.csv", row.names=1, header=TRUE)

#exploring structure and dimensions of dataset
str(humanW)

dim(humanW)
```

### Discriptive statistics summaries

Basic summaries of the data variables

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(knitr)
library(kableExtra)
```


```{r}
Label<-c("Edu2.FM" ,     "Labo.FM",       "Edu.Exp"    ,   "Life.Exp",
 "GNI"     ,      "Mat.Mor"  ,    "Ado.Birth", "Parli.F"  )

Variable<-c("Population with at least some secondary education (female/male ratio)",
 "Labour force participation rate (female/male ratio)",
"Expected years of schooling",
"Life expectancy at birth",
"Gross national income (GNI) per capita",
"Maternal mortality ratio(deaths per 100 000 live births)",
"Adolescent birth rate (births per 1 000 women ages 15-19)",
"Share of seats in parliament (female)")

om<- data.frame(Label,Variable)

kable(om, "html") %>%
  kable_styling(bootstrap_options = "bordered", full_width = TRUE)
```


```{r}
summary(humanW)
```

More summary statistics and boxplots 

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(settings)
library(tableone)
```


```{r}
reset(options)
options("scipen"=10, "digits"=2)
tab1<-CreateTableOne(vars=c( "Edu2.FM" ,"Labo.FM" ,  "Edu.Exp"   ,   
"Life.Exp" ,     "GNI"  ,         "Mat.Mor"  ,    "Ado.Birth",
"Parli.F") ,data=humanW)
summary(tab1)

```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
reset(options)

```

```{r}
par(mfrow=c(2,4))

p1<-boxplot(humanW$Edu2.FM,col="lightgreen",main="F/M ratio of educational level",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p2<-boxplot(humanW$Labo.FM,col="brown",main="F/M ratio labour force",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p3<-boxplot(humanW$Edu.Exp,col="lightblue",main="Expected years of schooling",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p4<-boxplot(humanW$Life.Exp,col="green",main="Life expectancy",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p5<-boxplot(humanW$GNI,col="orange",main="National income percap",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p6<-boxplot(humanW$Mat.Mor,col="grey",main="Maternal mortality (per 100000)",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p7<-boxplot(humanW$Ado.Birth,col="purple",main="Adolescent birth rate",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p8<-boxplot(humanW$Parli.F,col="yellow",main="Share of female seats in parliament",cex.main=0.9, cex=1, pch=1, pt.cex =1)

```

If we now try to have a look on African countries
```{r}
# Africa comprises all african counties from 
# https://www.countries-ofthe-world.com/countries-of-africa.html

Africa <- c("Algeria", "Angola", "Benin", "Botswana", "Burkina Faso", "Burundi", "Cabo Verde", "Cameroon", "Central African Republic (CAR)", "Chad", "Comoros", "Democratic Republic of the Congo", "Republic of the Congo", "Cote d'Ivoire", "Djibouti", "Egypt", "Equatorial Guinea", "Eritrea", "Ethiopia", "Gabon", "Gambia", "Ghana", "Guinea", "Guinea-Bissau", "Kenya", "Lesotho", "Liberia", "Libya", "Madagascar", "Malawi", "Mali", "Mauritania", "Mauritius", "Morocco", "Mozambique", "Namibia", "Niger", "Nigeria", "Rwanda", "Sao Tome and Principe", "Senegal", "Seychelles", "Sierra Leone", "Somalia", "South Africa", "South Sudan", "Sudan", "Swaziland", "Tanzania", "Togo", "Tunisia", "Uganda", "Zambia", "Zimbabwe")

# intersect Africa with row names return existing African countries

human_Africa <- humanW[intersect(Africa, row.names(humanW)),]

```


```{r}
reset(options)
options("scipen"=10, "digits"=2)
tab1<-CreateTableOne(vars=c( "Edu2.FM" ,"Labo.FM" ,  "Edu.Exp"   ,   
"Life.Exp" ,     "GNI"  ,         "Mat.Mor"  ,    "Ado.Birth",
"Parli.F") ,data=human_Africa)
summary(tab1)

```




```{r}
par(mfrow=c(2,4))

p1<-boxplot(human_Africa$Edu2.FM,col="lightgreen",main="F/M ratio of educational level",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p2<-boxplot(human_Africa$Labo.FM,col="brown",main="F/M ratio labour force",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p3<-boxplot(human_Africa$Edu.Exp,col="lightblue",main="Expected years of schooling",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p4<-boxplot(human_Africa$Life.Exp,col="green",main="Life expectancy",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p5<-boxplot(human_Africa$GNI,col="orange",main="National income percap",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p6<-boxplot(human_Africa$Mat.Mor,col="grey",main="Maternal mortality (per 100000)",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p7<-boxplot(human_Africa$Ado.Birth,col="purple",main="Adolescent birth rate",cex.main=0.9, cex=1, pch=1, pt.cex =1)

p8<-boxplot(human_Africa$Parli.F,col="yellow",main="Share of female seats in parliament",cex.main=0.9, cex=1, pch=1, pt.cex =1)

```






















